Placing the spotted T Tauri star LkCa 4 on an HR diagram

### 2016-08-16 document/ms.tex
Almost done with version 0.6... will send to co-authors soon.

\keywords{stars: fundamental parameters --- stars: individual (\name) ---  stars: low-mass -- stars: statistics}

@@

We took two approaches to characterizing the effective temperature response from a spotted-star spectrum.  First we measured a single effective temperature to unique spectral orders in the optical (\S \ref{sec:ESP_starfish}) and near-IR (\S \ref{sec:IGR_starfish}).  Second we fit a two-temperature mixture model (\S \ref{sec:methods}) to the near-IR spectra (\S \ref{sec:two_tempIGRINS}), which should show greater starspot flux density than optical(\S \ref{sec:whyNearIR}).

@@

The heightened sensitivity to starspot spectral lines as a function of wavelength can be understood in the following way.  Starspots are cooler than their surrounding photosphere and will, therefore, have a longer wavelength of peak emission, according to Wein's displacement law.  The average ratio of flux density between starspot and bulk photosphere will increase with wavelength until asymptoting to a fixed value in the Rayleigh-Jeans tail \citep{wolk96}.

@@

\begin{table}
\caption{Predicted Photometric Range}
\label{tab:photrange}
\begin{tabular}{cccc}
\hline
Band & Low Spot & High Spot & Range\\
Cool Spot & 68\% & 83\% &\\
\hline
\multicolumn{4}{c}{Based on $V$, need to adapt for $A_V$}\\
$B$ & 13.80 & 14.48 & 0.68\\
$V$ & 12.56 & 13.23 & 0.67\\
$R$ & 11.53 & 12.17 & 0.64 \\
$I$ & 10.80 & 11.38 & 0.58\\
$J$ & 9.79 & 10.00 & 0.21\\
$K$ & 8.89 & 9.04 & 0.15\\
\hline
\end{tabular}
\end{table}

@@

Some overview?

@@

\comm{From discussion with GJH about lifetime of spots:  Where to put the source on the observational HR diagram depends on what your assumptions for its evolutionary history are.}

{\bf goes later?}

Despite some evidence that LkCa 4 is an outlier, spots are likely broadly important in estimating the properties of young low-mass stars. As an example, the well-studied CTTS TW Hya has a $\sim 4000$ K photosphere measured from high resolution optical spectra \citep[e.g.][]{yang05}.  However, the near-IR spectrum is consistent with $\sim M2$, when compared with main sequence dwarf stars \citep{vacca11}.  \citet{debes13} reconciled this discrepancy by invoking spots, although the spectral contribution from continuum emission was uncertain.  \citet{mcclure13} found that the TW Hya spectrum was well reproduced with near-IR spectra of stars with similar ($\sim$K7) optical spectral type, as long as the comparison spectra were young.   \citet{herczeg14} found an intermediate spectral type for TW Hya and other TTSs by focusing on TiO bands at redder wavelengths than previous spectra, and also by using young stars as photospheric templates.  The mismatch between near-IR spectra of dwarf stars and young stars, even with similar optical spectral types, is likely explained by the prevalence and importance of cool spots in low mass stars.

@@

The spectral inference method assumed only two temperature components present in the photosphere.  In reality, starspots are probably described by a range of effective temperatures like umbra, penumbra, and plages on the sun's surface.  A model possessing three unique temperatures and two unique fill factors might better approximate the minor constituents of the emergent spectrum than our two temperature model can, but would always fit the data better.  The fit-quality of the existing two-temperature fits seems fairly strong through much of $H-$band and many $K-$band orders, so the contribution of minor photospheric constituents must be relatively low.  Interestingly, $K-$band has several spectral orders that converge towards hybrid temperatures in the vicinity of $\sim3500$ K.  Most of these orders were flagged as erroneous due to strong metal line spectral outliers or uncorrected calibration artifacts.  But some orders just showed mediocre fits where no permutation of hot and cool two-temperature models could describe all of the variance in the spectrum.  These $K-$band orders could be examined for a penumbra-like temperature component.  We suspect systematic errors in the models would ultimately limit the interpretability of models with more than two temperature components.  High-precision polychromatic planetary transit photometry could offer an avenue to measure the morphologies and temperatures of starspots for the subset of transiting exoplanet host stars.

We assumed that the \PHOENIX\ synthetic spectra are a good representation of the observed IGRINS spectrum of LkCa 4, a young star that possesses a strong magnetic field \citep{donati14}.  Cool sunspots are associated with heightened magnetic fields on the sun.  Magnetic fields will have two effects on the emergent spectrum- line broadening attributable to the Zeeman effect, and pressure broadening attributable to seeing deeper into deeper layers in the photosphere compared to the surrounding hotter photosphere.  The former effect will differentially impact lines with high magnetic sensitivity \citep{deen13}.  The latter effect could masquerade as an apparent shift in surface gravity, as seen in mixture models of spotted stars with empirical templates composed of dwarfs and giants \citep{oneal96}.  Our tacit assumptions have been that these effects are 1) relatively small compared to the gross appearance and disappearance of spectral lines at disparate temperatures, 2) there exist many orders and lines that are not affected at all, and 3) the Gaussian process noise model fosters some resilience against unknown residual spectrum correlations.  We see tentative evidence for physics outside of the pre-computed model grids in the poor fits to the observed spectra in $K-$band.  Strong metal lines demonstrate large departures in their best fit $\Z$ due to Zeeman broadening.  No pre-computed model grids including a range of surface magnetic fields are available at the wavelength range and spectral sampling needed to approach these questions within this spectral inference framework.

Uncertainties in the \PHOENIX\ synthetic spectra continuum opacity flow down to uncertainties in the absolute value of the fill factor.  Erroneous bulk opacities could be responsible for some of the scatter of fill factor distributions seen across different spectral orders, as in Figures \ref{fig:TwoTempResults} and \ref{fig:specPostageStamp}.  The large number of spectral orders over which the fill factors are averaged offers some resilience against these anomalies.

@@

\subsection{Is \name an extreme source?}
LkCa 4 is an extreme source by several metrics.  Mainly, its large amplitude of cyclical optical variability exceeds all other counterparts from long-term monitoring of WTTSs \citep{grankin08}.  LkCa 4 has about 40 times the photometric amplitude of the typical young star towards Ophiuchus or Upper Sco (Gully-Santiago et al. in prep.).

@@

@@

Table \ref{tbl_order_results} lists all the IGRINS spectral order results.
\comm{Update this with ``used/unused'' label and most recent values.}
\input{tables/tbl_order_results.tex}

@@

This research has made use of NASA's Astrophysics Data System.  
Based on data from the OMC Archive at CAB (INTA-CSIC), pre-processed by ISDC.
We acknowledge with thanks the variable star observations from the AAVSO International Database contributed by observers worldwide and used in this research.
This publication makes use of data products from the Two Micron All Sky Survey, which is a joint project of the University of Massachusetts and the Infrared Processing and Analysis Center/California Institute of Technology, funded by the National Aeronautics and Space Administration and the National Science Foundation.

>>>>

\keywords{stars: fundamental parameters --- stars: individual (LkCa 4) ---  stars: low-mass -- stars: statistics}

@@

We took two approaches to characterizing the effective temperature response from a spotted-star spectrum.  First we measured a single effective temperature to unique spectral orders in the optical (\S \ref{sec:ESP_starfish}) and near-IR (\S \ref{sec:IGR_starfish}).  Second we fit a two-temperature mixture model (\S \ref{sec:methods}) to a high grasp near-IR spectrum (\S \ref{sec:two_tempIGRINS}), which should show greater starspot flux density than optical.

@@

The heightened sensitivity to starspot spectral lines as a function of wavelength can be understood in the following way.  Starspots are cooler than their surrounding photosphere and will, therefore, have a longer wavelength of peak emission, according to Wein's displacement law.  The average ratio of flux density between starspot and bulk photosphere will increase with wavelength until asymptoting to a fixed value in the Rayleigh-Jeans tail \citep{wolk96}.  The bottom panel of Figure \ref{fig:SingleTeffvsOrder} shows the flux density ratio for patches of photosphere with equal areas (50\% filling factor) but different temperatures-- 2800 K and 4100 K for the starspot and ambient photosphere respectively.  The black body ratios predict smooth flux ratio increases with wavelength, while the ratio of \PHOENIX\ models shows wavelength regions with heightened sensitivity to starspot fluxes, for example in $J-$band.

@@

\begin{table}
\caption{Predicted Photometric Range}
\centering
\label{tab:photrange}
\begin{tabular}{cccc}
\hline
Band & Low Spot & High Spot & Range\\
Cool Spot & 68\% & 83\% &\\
\hline
\multicolumn{4}{c}{Based on $V$, need to adapt for $A_V$}\\
$B$ & 13.80 & 14.48 & 0.68\\
$V$ & 12.56 & 13.23 & 0.67\\
$R$ & 11.53 & 12.17 & 0.64 \\
$I$ & 10.80 & 11.38 & 0.58\\
$J$ & 9.79 & 10.00 & 0.21\\
$K$ & 8.89 & 9.04 & 0.15\\
\hline
\end{tabular}
\end{table}

@@

\todo{Some overview?}

@@

As seen in Figures \ref{fig:PhotTime} and \ref{fig:PhotPhase}, the areal coverage of spots appears to wax and wane secularly through the observing seasons of the last 31 years.  It is conceivable that LkCa 4 happens to be going through a short-lived episode of relatively high coverage of spots analogous to solar maximum.  The intensity and duration of such episodes would further confound the interpretation of its instantaneous position on the HR diagram.

@@

The spectral inference method assumed only two temperature components present in the photosphere.  In reality, starspots are probably described by a range of effective temperatures like umbra, penumbra, and plages on the sun's surface.  A model possessing three unique temperatures and two unique fill factors might better approximate the minor constituents of the emergent spectrum than our two temperature model can, but would always fit the data better and could lead to the possibility of overfitting.  The fit-quality of the existing two-temperature fits seems fairly strong through much of $H-$band and many $K-$band orders, so the contribution of minor photospheric constituents must be relatively low.  Interestingly, $K-$band has several spectral orders that converge towards hybrid temperatures in the vicinity of $\sim3500$ K.  Most of these orders were flagged as erroneous due to strong metal line spectral outliers or uncorrected calibration artifacts.  But some orders just showed mediocre fits where no permutation of hot and cool two-temperature models could describe all of the variance in the spectrum.  These $K-$band orders could be examined for a penumbra-like temperature component.  We caution that systematic errors in the models would ultimately limit the interpretability of models with more than two temperature components.  Space-based high-precision planetary transit spectrophotometry could offer an avenue to measure the morphologies and temperature distributions of starspots for the subset of transiting exoplanet host stars or eclipsing binaries.

We assumed that the \PHOENIX\ synthetic spectra are a good representation of the observed IGRINS spectrum of LkCa 4, a young star that possesses a strong magnetic field \citep{donati14}.  Cool sunspots are associated with locally heightened magnetic fields on the sun, so synthetic spectral models employing negligible magnetic fields will fail to fit magnetic sensitive features.  Magnetic fields will have two effects on the emergent spectrum- line broadening attributable to the Zeeman effect, and pressure broadening attributable to seeing deeper into deeper layers in the photosphere compared to the surrounding hotter photosphere.  The former effect will differentially impact lines with high magnetic sensitivity \citep[\emph{e.g.}][]{johnskrull99,deen13}.  The latter effect could masquerade as an apparent shift in surface gravity, as seen in mixture models of spotted stars with empirical templates composed of dwarfs and giants \citep{oneal96}.  Our tacit assumptions have been that these line-broadening effects are 1) relatively small compared to the gross appearance and disappearance of spectral lines at disparate temperatures, 2) there exist many orders and lines that are not affected at all, and 3) the Gaussian process noise model fosters some resilience against unknown residual spectrum correlations induced by such under-fitting.  We see tentative evidence for physics outside of the pre-computed model grids in the poor fits to the observed spectra in $K-$band.  Strong metal lines demonstrate large departures in their best fit $\Z$, as the na\:ive fitting strategy futilely attempts to minimize the residuals from Zeeman-broadened lines.  No pre-computed model grids including a range of surface magnetic fields are available at the wavelength range and spectral sampling needed to approach these questions within this spectral inference framework.

Finally, uncertainties in the \PHOENIX\ synthetic spectra continuum opacity flow down to uncertainties in the absolute value of the fill factor.  Erroneous bulk opacities could be responsible for some of the scatter of fill factor distributions seen across different spectral orders, as in Figures \ref{fig:TwoTempResults} and \ref{fig:specPostageStamp}.  The large number of spectral orders over which the fill factors are averaged offers resilience against these anomalies.

@@

\subsection{Is \name an extreme source?}
LkCa 4's large amplitude of cyclical optical variability exceeds all other counterparts from long-term monitoring of WTTSs \citep{grankin08}.  From a statistical perspective, it might not be a surprise that the source with the largest photometric variability also posessess among the largest coverage fraction of cool spots.  The large $\sim$1 kG toroidal component at equatorial latitudes makes LkCa 4 an outlier by this metric, when compared to other TTSs.

Despite the evidence that LkCa 4 is an outlier, spots are likely broadly important in estimating the properties of young low-mass stars. As an example, the well-studied CTTS TW Hya has a $\sim 4000$ K photosphere measured from high resolution optical spectra \citep[e.g.][]{yang05}.  However, the near-IR spectrum is consistent with $\sim M2$, when compared with main sequence dwarf stars \citep{vacca11}.  \citet{debes13} reconciled this discrepancy by invoking spots, although the spectral contribution from continuum emission was uncertain.  \citet{mcclure13} found that the TW Hya spectrum was well reproduced with near-IR spectra of stars with similar ($\sim$K7) optical spectral type, as long as the comparison spectra were young.   \citet{herczeg14} found an intermediate spectral type for TW Hya and other TTSs by focusing on TiO bands at redder wavelengths than previous spectra, and also by using young stars as photospheric templates.  The mismatch between near-IR spectra of dwarf stars and young stars, even with similar optical spectral types, is likely explained by the prevalence and importance of cool spots in low mass stars.

@@

Table \ref{tbl_order_results} lists all the IGRINS spectral order results.
\LongTables
\input{tables/tbl_order_results.tex}

@@

@@

This work used the Immersion Grating Infrared Spectrograph (IGRINS) that was developed under a collaboration between the University of Texas at Austin and the Korea Astronomy and Space Science Institute (KASI) with the financial support of the US National Science Foundation under grant ASTR1229522, of the University of Texas at Austin, and of the Korean GMT Project of KASI.

This research has made use of NASA's Astrophysics Data System.  The research is based on data from the OMC Archive at CAB (INTA-CSIC), pre-processed by ISDC.
We acknowledge with thanks the variable star observations from the AAVSO International Database contributed by observers worldwide and used in this research.
This publication makes use of data products from the Two Micron All Sky Survey, which is a joint project of the University of Massachusetts and the Infrared Processing and Analysis Center/California Institute of Technology, funded by the National Aeronautics and Space Administration and the National Science Foundation.
